---
title: "Spatio-Temporal Analysis of HIV and Tuberculosis Co-Morbidity in Kenya"
subtitle: "County-Level Bayesian Joint Disease Modelling, 2012–2017"
author: "Verrah A. Otiende"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: references.bib
---

---

# Overview

This document presents a fully reproducible Bayesian spatio-temporal **joint
disease model** for HIV and Tuberculosis (TB) incidence across Kenya's **47
counties** from **2012 to 2017**. The analysis is based on case notification
data obtained from the Kenya Ministry of Health and population projections from
the Kenya National Bureau of Statistics (KNBS).

The analysis proceeds through six stages:

1. **Package loading** — load all required R libraries
2. **Data import** — read raw case counts and population projections from CSV files
3. **Indirect standardisation** — compute Standardised Incidence Ratios (SIRs)
   at county level for each year and disease
4. **Spatial data preparation** — merge tabular data with county shapefiles and
   produce descriptive SIR choropleth maps
5. **INLA model fitting** — fit a shared-component Bayesian spatio-temporal
   Poisson model using R-INLA
6. **Results extraction and visualisation** — extract posterior estimates of
   spatial effects, temporal trends, and space-time interaction probabilities,
   and display them as maps and line plots

---

# Package Dependencies

**What this chunk does:**
All R packages required for the entire analysis are loaded in a single block at
the top of the script. This is deliberate good practice — it makes dependencies
immediately visible to any reader and ensures the workspace is fully set up
before any analysis code runs. The packages cover four areas: spatial data
handling, mapping and visualisation, epidemiological tools, and Bayesian
modelling.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,    # Show all code in the rendered output
  message   = FALSE,   # Suppress package startup messages
  warning   = FALSE,   # Suppress warnings in the rendered output
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)
```

```{r packages}
# ── Spatial data handling ─────────────────────────────────────────────────────
library(sp)           # Core spatial classes: SpatialPolygonsDataFrame, etc.
library(sf)           # Modern simple-features spatial objects
library(rgdal)        # Read and write OGC spatial formats including .shp
library(spdep)        # Spatial weights, neighbourhood lists, Moran tests
library(maptools)     # Utilities for handling and manipulating spatial objects
library(spacetime)    # Classes and methods for spatio-temporal data

# ── Mapping and visualisation ─────────────────────────────────────────────────
library(maps)         # Basic map-drawing functions
library(mapdata)      # High-resolution geographic boundary data
library(RColorBrewer) # Carefully chosen colour palettes for maps and charts
library(lattice)      # Trellis-style panel graphics (required by spplot)
library(latticeExtra) # Additional panel functions extending lattice
library(gridExtra)    # Combine multiple lattice panels into a single grid
library(grid)         # Low-level grid graphics primitives
library(car)          # Miscellaneous regression and plot utilities

# ── Epidemiological and statistical tools ─────────────────────────────────────
library(epitools)     # Epidemiological rate and ratio calculations
library(DCluster)     # Spatial cluster detection for disease data
library(stats)        # Base R statistical functions
library(plotrix)      # Specialised plot types and diagram utilities

# ── Bayesian spatio-temporal modelling ───────────────────────────────────────
library(INLA)         # Integrated Nested Laplace Approximation (R-INLA)

# ── Reporting ─────────────────────────────────────────────────────────────────
library(knitr)        # Dynamic report generation and table formatting
```

**Output:** No visible output is produced by this chunk. If any package is not
installed, R will throw an error here and stop. Install missing packages with
`install.packages("package_name")`. INLA requires a separate step:
```r
install.packages("INLA",
  repos = c(getOption("repos"),
            INLA = "https://inla.r-inla-download.org/R/stable"),
  dep = TRUE)
```

---

# Data Import

## File Paths

**What this chunk does:**
Rather than hard-coding file paths throughout the script (which would require
changing them in many places when moving to a new machine), all three file
paths are defined here in one place. Update only this block when running the
analysis on a different machine. `PATH_HIV` and `PATH_TB` point to the CSV
data files; `PATH_SHP` points to the Kenya county polygon shapefile.

```{r file-paths}
PATH_HIV <- "data/HIV_Data_2018.csv"          # HIV case counts and populations
PATH_TB  <- "data/TB_Data_2018.csv"           # TB case counts and populations
PATH_SHP <- "data/County_Kenya/County.shp"    # Kenya county boundary shapefile
```

**Output:** No visible output. These are character string variables stored in
memory and referenced by all downstream chunks.

---

## Load Case Count Data

**What this chunk does:**
The two CSV files are read into R as data frames. Each file has 47 rows (one
per county) and 16 columns: a county identifier, a county name, six years of
case counts (2012–2017), six corresponding projected populations, and two
pooled totals. A `stopifnot()` check immediately confirms that both files
contain exactly 47 rows — if either file is missing rows, the analysis stops
here with an informative error rather than producing silent wrong results.
The `cat()` statements print a quick confirmation of the dimensions loaded.

```{r load-data}
# Read HIV new-diagnosis counts and projected populations (47 counties x 16 cols)
HIV_data <- read.csv(PATH_HIV)

# Read TB new-notification counts and projected populations (same structure)
TB_data  <- read.csv(PATH_TB)

# Validation: stop immediately if either dataset does not have exactly 47 rows
stopifnot(nrow(HIV_data) == 47, nrow(TB_data) == 47)

# Confirm successful loading
cat("HIV dataset:", nrow(HIV_data), "counties,", ncol(HIV_data), "columns\n")
cat("TB  dataset:", nrow(TB_data),  "counties,", ncol(TB_data),  "columns\n")
```

**Output:** Two confirmation lines, e.g.:
```
HIV dataset: 47 counties, 16 columns
TB  dataset: 47 counties, 16 columns
```
If the `stopifnot()` fails, R will print an error and halt — this is
intentional to prevent the rest of the analysis from running on incomplete data.

---

# Indirect Standardisation and SIR Computation

County populations differ enormously across Kenya (Nairobi: >3.5 million;
Lamu: <115,000). A raw count of cases therefore cannot fairly compare disease
burden between counties. **Indirect standardisation** adjusts for population
size differences by computing how many cases *would be expected* in each county
if its residents experienced the national average incidence rate.

The **Standardised Incidence Ratio (SIR)** is the key summary measure:

$$
\text{SIR}_{ic} = \frac{O_{ic}}{E_{ic}}, \qquad
E_{ic} = \bar{\lambda}_c \times \bar{P}_i
$$

where:

- $O_{ic}$ = observed cases in county $i$ during period $c$
- $E_{ic}$ = expected cases (if county $i$ had the national rate $\bar{\lambda}_c$)
- $\bar{\lambda}_c$ = national crude rate for period $c$ (total cases ÷ total population)
- $\bar{P}_i = P_{i,\text{Total}} / 6$ = county $i$'s average annual population

An SIR > 1 indicates above-national-average burden; SIR < 1 indicates
below-average burden; SIR = 1 means exactly at the national average.

---

## Standard Populations

**What this chunk does:**
The *standard population* for each county is its average annual projected
population across the six study years. This is computed by dividing the
six-year total population (`PP_Total`) by 6. Using the average rather than any
single year's population ensures the denominator is stable and not unduly
influenced by year-to-year fluctuations. The result is a 47-element numeric
vector, one value per county.

```{r standard-populations}
# Average annual projected population over the 6-year study period (per county)
standard.HIV <- HIV_data$PP_Total / 6   # 47-element vector from the HIV file
standard.TB  <- TB_data$PP_Total  / 6   # 47-element vector from the TB file
```

**Output:** No visible output. Two numeric vectors of length 47. Note that
`standard.HIV` and `standard.TB` will be numerically identical because both CSV
files use the same KNBS population projections; they are kept as separate objects
for clarity and to allow future modification.

---

## National Crude Incidence Rates

**What this chunk does:**
The national crude rate for each year is total new cases across all 47 counties
divided by the total projected population across all 47 counties. Rather than
writing a separate line for each of the six years (12 repetitive lines per
disease), `sapply()` loops over the years vector and computes all six rates in
one call. The result is a named numeric vector. A pooled rate across all six
years is also computed. A formatted table displays the rates per 100,000
population so they are interpretable at a glance.

```{r crude-rates}
years <- 2012:2017   # Study period as an integer vector

# ── HIV national crude rates ──────────────────────────────────────────────────
# For each year: sum all HIV cases / sum all projected populations (47 counties)
cruderate.HIV <- sapply(years, function(y) {
  sum(HIV_data[[paste0("HIV_", y)]]) / sum(HIV_data[[paste0("PP_", y)]])
})
names(cruderate.HIV) <- years

# Pooled 6-year rate: total cases across all years / total person-years
cruderate.HIV_Total <- sum(HIV_data$HIV_Total) / sum(HIV_data$PP_Total)

# ── TB national crude rates ───────────────────────────────────────────────────
cruderate.TB <- sapply(years, function(y) {
  sum(TB_data[[paste0("TB_", y)]]) / sum(TB_data[[paste0("PP_", y)]])
})
names(cruderate.TB) <- years

cruderate.TB_Total <- sum(TB_data$TB_Total) / sum(TB_data$PP_Total)

# ── Display as a formatted table (scaled to per 100,000 for readability) ─────
crude_summary <- data.frame(
  Year                    = years,
  `HIV Rate (per 100,000)` = round(cruderate.HIV * 100000, 2),
  `TB Rate (per 100,000)`  = round(cruderate.TB  * 100000, 2)
)
knitr::kable(crude_summary,
             caption = "Table 1. National annual crude incidence rates for HIV and TB, 2012-2017.")
```

**Output:** A formatted table with three columns (Year, HIV rate, TB rate).
These rates are the national benchmarks against which every county's SIR is
calculated — a county with SIR > 1 has an incidence rate above the
corresponding value in this table. Comparing the two rate columns across years
also gives a first impression of whether national incidence is rising, falling,
or stable over the study period, and whether HIV and TB track each other over
time.

---

## Expected Case Counts

**What this chunk does:**
The expected count for each county-year is calculated by multiplying the
national crude rate by the county's standard population. This answers: "How
many cases would we expect in this county if its residents experienced the
national rate?" `sapply()` loops over the six years to produce a 47 × 6 matrix
for each disease (counties in rows, years in columns). A separate pooled
expected count (using the 6-year rate) is also computed for each county.
These expected counts serve as the **Poisson offset** (`E` argument) when the
INLA model is fitted in Section 6.

```{r expected-counts}
# ── HIV expected counts: national crude rate x county standard population ─────
# Result: 47 x 6 matrix — each cell = expected HIV cases for county i in year y
expected.HIV <- sapply(years, function(y) {
  cruderate.HIV[as.character(y)] * standard.HIV
})
colnames(expected.HIV) <- years

# Pooled expected count using the 6-year overall crude rate
expected.HIV_Total <- cruderate.HIV_Total * standard.HIV

# ── TB expected counts ────────────────────────────────────────────────────────
expected.TB <- sapply(years, function(y) {
  cruderate.TB[as.character(y)] * standard.TB
})
colnames(expected.TB) <- years

expected.TB_Total <- cruderate.TB_Total * standard.TB
```

**Output:** No visible output. Four objects are created: `expected.HIV` and
`expected.TB` are 47 × 6 numeric matrices; `expected.HIV_Total` and
`expected.TB_Total` are numeric vectors of length 47. Counties with large
standard populations (e.g. Nairobi, Kakamega) will have large expected counts;
small, sparsely populated counties (e.g. Lamu, Isiolo) will have small expected
counts, making the SIR a fair population-adjusted comparison.

---

## SIR Computation (Continuous)

**What this chunk does:**
The SIR for each county-year is computed by dividing the observed count by the
corresponding expected count. `sapply()` again loops over years to produce a
47 × 6 matrix. A pooled SIR using 6-year totals is also computed per county.
Counties where observed > expected (SIR > 1) have above-average burden after
adjusting for population size; counties where observed < expected (SIR < 1)
are doing better than the national average.

```{r sir-continuous}
# ── HIV SIRs: observed / expected (element-wise) ─────────────────────────────
adj.HIV <- sapply(years, function(y) {
  HIV_data[[paste0("HIV_", y)]] / expected.HIV[, as.character(y)]
})
colnames(adj.HIV) <- years

# Pooled HIV SIR (using 6-year totals)
adj.HIV.total <- HIV_data$HIV_Total / expected.HIV_Total

# ── TB SIRs ───────────────────────────────────────────────────────────────────
adj.TB <- sapply(years, function(y) {
  TB_data[[paste0("TB_", y)]] / expected.TB[, as.character(y)]
})
colnames(adj.TB) <- years

adj.TB.total <- TB_data$TB_Total / expected.TB_Total
```

**Output:** No visible output. `adj.HIV` and `adj.TB` are 47 × 6 numeric
matrices. Values typically range from near zero (very low burden counties) to
well above 1.0 (high-burden counties). These raw SIRs are used directly as
inputs to the INLA model; the choropleth maps in Section 5 display their
categorical versions.

---

## SIR Classification (Categorical)

**What this chunk does:**
For choropleth mapping, continuous SIRs are grouped into four ordered risk
categories. A single set of breakpoints is applied consistently across all years
and both diseases, so the colour scale is directly comparable across every map
in the figure. A reusable helper function `classify_sir()` wraps the `cut()`
call so it can be applied to any SIR vector in one line. `apply()` then applies
it column-by-column (i.e. year by year) across the full SIR matrix.

```{r sir-categorical}
# Breakpoints: 0.40, 0.70, 0.90 are the internal boundaries; 6 catches all high values
sir.breaks <- c(0, 0.40, 0.70, 0.90, 6)
sir.labels <- c("0.00-0.40", "0.41-0.70", "0.71-0.99", "1.00+")

# Reusable helper: convert a continuous SIR vector to a four-level factor
classify_sir <- function(x) {
  cut(x, breaks = sir.breaks, include.lowest = TRUE, labels = sir.labels)
}

# ── Apply to HIV SIR matrix (each column = one year) ─────────────────────────
adj_HIV_cat       <- apply(adj.HIV, 2, classify_sir)
colnames(adj_HIV_cat) <- paste0("adj_HIV_", years)
adj_HIV_total_cat <- classify_sir(adj.HIV.total)

# ── Apply to TB SIR matrix ────────────────────────────────────────────────────
adj_TB_cat        <- apply(adj.TB, 2, classify_sir)
colnames(adj_TB_cat) <- paste0("adj_TB_", years)
adj_TB_total_cat  <- classify_sir(adj.TB.total)
```

**Output:** No visible output. Factor variables are created with four levels:
"0.00–0.40" (very low: county rate < 40% of national average), "0.41–0.70"
(low), "0.71–0.99" (approaching national average), and "1.00+" (at or above
national average). These factors drive the colour coding in all SIR maps.

---

## Build the Consolidated Analysis Dataset

**What this chunk does:**
All computed variables (observed counts, expected counts, continuous SIRs, and
categorical SIRs) are assembled into a single wide data frame called
`data_prop`. Each row represents one county; columns cover all six years and
both diseases. The `setNames()` calls give each group of columns a consistent
naming convention (e.g. `obs.HIV2012`, `exp.TB2015`, `adj_HIV_2013`). This
master data frame is then: (1) merged with the shapefile for mapping, and (2)
reshaped into long format for INLA. The final `cat()` reports the dimensions
as a sanity check.

```{r build-data-prop}
# ── Observed counts: one column per year per disease ─────────────────────────
obs_HIV <- setNames(as.data.frame(
  sapply(years, function(y) HIV_data[[paste0("HIV_", y)]])),
  paste0("obs.HIV", years))

obs_TB  <- setNames(as.data.frame(
  sapply(years, function(y) TB_data[[paste0("TB_", y)]])),
  paste0("obs.TB", years))

# ── Expected counts ───────────────────────────────────────────────────────────
exp_HIV <- setNames(as.data.frame(expected.HIV), paste0("exp.HIV", years))
exp_TB  <- setNames(as.data.frame(expected.TB),  paste0("exp.TB",  years))

# ── Continuous SIRs ───────────────────────────────────────────────────────────
sir_HIV <- setNames(as.data.frame(adj.HIV), paste0("adj.HIV.", years))
sir_TB  <- setNames(as.data.frame(adj.TB),  paste0("adj.TB.",  years))

# ── Categorical SIRs ──────────────────────────────────────────────────────────
cat_HIV <- setNames(as.data.frame(adj_HIV_cat), paste0("adj_HIV_", years))
cat_TB  <- setNames(as.data.frame(adj_TB_cat),  paste0("adj_TB_",  years))

# ── Assemble into one master data frame (47 rows = 47 counties) ───────────────
data_prop <- data.frame(
  County.ID     = HIV_data$Sn.no,
  County        = HIV_data$County,
  obs_HIV,
  obs.HIVtotal  = HIV_data$HIV_Total,
  obs_TB,
  obs.TBtotal   = TB_data$TB_Total,
  exp_HIV,
  exp.HIVtotal  = expected.HIV_Total,
  exp_TB,
  exp.TBtotal   = expected.TB_Total,
  sir_HIV,
  adj.HIV.total,
  sir_TB,
  adj.TB.total,
  cat_HIV,
  adj_HIV_total_cat,
  cat_TB,
  adj_TB_total_cat
)

cat("data_prop dimensions:", nrow(data_prop), "rows x", ncol(data_prop), "columns\n")
```

**Output:** A single dimension-confirmation line, e.g.:
`data_prop dimensions: 47 rows x 60 columns`. The 47 rows are the 47 counties.
The many columns arise because each disease × year combination contributes
observed, expected, continuous SIR, and categorical SIR columns. This
wide-format structure is used for choropleth mapping in Section 5.

---

# Spatial Data Preparation

## Load and Align the Shapefile

**What this chunk does:**
The Kenya county polygon shapefile is read into R as a `SpatialPolygonsDataFrame`
using `readOGR()`. Its attribute table contains a field `COUNTY` with county
names. These county names may not appear in the same order as the rows in
`data_prop` (which comes from the CSV files sorted alphabetically). The
`match()` call finds — for each county name in the shapefile — the
corresponding row index in `data_prop`. This reordering step is critical: if
the rows are not aligned before merging, every SIR value will be assigned to
the wrong county polygon on the map. A warning is issued for any county whose
name does not match.

```{r load-shapefile}
# Read Kenya county polygons
kenya <- readOGR(PATH_SHP, verbose = FALSE)

# Reorder data_prop rows to match the shapefile's internal county ordering
shp_order <- match(kenya$COUNTY, data_prop$County)
data_pop   <- data_prop[shp_order, ]

# Report successful matches
cat("Counties successfully matched:", sum(!is.na(shp_order)), "/ 47\n")

# Warn if any county name in the shapefile has no match in the CSV
if (any(is.na(shp_order))) {
  warning("Unmatched counties: ",
          paste(kenya$COUNTY[is.na(shp_order)], collapse = ", "))
}
```

**Output:** A confirmation line such as `Counties successfully matched: 47 / 47`.
If fewer than 47 match, the warning identifies which county names differ between
the shapefile and the CSV (e.g. spelling differences, diacritics, or hyphenation).
Any mismatches must be resolved — typically by editing the CSV county name to
match the shapefile — before proceeding.

---

## Merge Shapefile with Attribute Data

**What this chunk does:**
`merge()` joins the spatial polygon object with the reordered analysis data
frame on the county name field. The output, `kenya.one`, is a
`SpatialPolygonsDataFrame` containing both the county boundary polygons and
all computed SIR, expected count, and observed count columns. The attribute
table is also extracted as a plain data frame (`data.kenya`) for non-spatial
operations later. This merged object is the foundation for all choropleth maps.

```{r merge-spatial}
# Join spatial polygons with the full analysis data frame on county name
kenya.one  <- merge(kenya, data_pop, by.x = "COUNTY", by.y = "County")

# Extract the attribute table (without geometry) for non-spatial use
data.kenya <- attr(kenya.one, "data")

cat("Merged spatial object:", nrow(data.kenya), "rows,",
    ncol(data.kenya), "columns\n")
```

**Output:** A confirmation such as `Merged spatial object: 47 rows, 62 columns`.
The 47 rows confirm all counties are present. At this point `kenya.one` is
ready for `spplot()` mapping — any column in the attribute table can be
visualised as a choropleth map by passing its name to `zcol`.

---

# Descriptive SIR Maps (Pre-Model)

These choropleth maps provide a purely descriptive spatial overview of HIV and
TB burden *before* the Bayesian model is applied. They display the raw
indirect-standardised SIR categories for each of the six study years, allowing
the reader to identify which counties have consistently high or low burden and
whether geographic patterns shift over time.

## Set Up the Colour Palette

**What this chunk does:**
`brewer.pal()` creates a four-colour Yellow-Orange-Red (YlOrRd) palette from
ColorBrewer. This palette is perceptually ordered — lightest yellow represents
the lowest SIR category (very low burden) and darkest red represents the
highest — making the maps immediately intuitive. The `trellis.par.set()` call
removes the default axis border lines that `spplot()` draws around each map
panel, producing a cleaner output.

```{r palette-setup}
# Four-colour Yellow-Orange-Red: low burden (yellow) to high burden (red)
my.palette.yor <- brewer.pal(4, "YlOrRd")

# Remove axis borders from spplot panels for a cleaner appearance
trellis.par.set(axis.line = list(col = NA))
```

**Output:** No visible output. `my.palette.yor` is a character vector of four
hex colour codes (e.g. `"#FFFFB2" "#FECC5C" "#FD8D3C" "#E31A1C"`). It is
passed to `col.regions` in every subsequent `spplot()` call.

---

## HIV SIR Maps (2012–2017)

**What this chunk does:**
`lapply()` loops over the six study years and creates one `spplot()` choropleth
panel for each year. Each panel shades the 47 counties according to their HIV
SIR category (`adj_HIV_YYYY`). `grid.arrange()` then lays the six panels into
a 2-row × 3-column grid, producing a single figure showing how the spatial
distribution of HIV burden evolved across the study period. The `cuts = 2`
argument matches the number of colour boundaries to the four categories defined
by three internal breakpoints.

```{r hiv-sir-maps, fig.cap="Figure 1. HIV Standardised Incidence Ratio (SIR) by county, 2012-2017. Yellow = very low (0.00-0.40); light orange = low (0.41-0.70); dark orange = moderate (0.71-0.99); red = high burden at or above national average (1.00+)."}
# Build one spplot panel per year, then combine into a 2x3 grid
hiv_panels <- lapply(years, function(y) {
  spplot(kenya.one,
         zcol         = paste0("adj_HIV_", y),
         col.regions  = my.palette.yor,
         colorkey     = list(space = "right", height = 1.0),
         cuts         = 2,
         main         = as.character(y),
         par.settings = list(fontsize = list(text = 12)))
})

do.call(grid.arrange, c(hiv_panels, list(nrow = 2, ncol = 3)))
```

**Output:** A 2 × 3 panel figure showing six HIV SIR maps (one per year).
Counties in red have HIV incidence at or above the national average; counties
in yellow are well below it. Consistent red shading across years for a given
county indicates a persistently high-burden area. Western Kenya counties (Homa
Bay, Kisumu, Siaya, Migori, Busia) and Nairobi are typically expected to
appear red given Kenya's known HIV epidemiology. Comparing the six panels
reveals whether geographic patterns are stable or shifting over time.

---

## TB SIR Maps (2012–2017)

**What this chunk does:**
Identical approach to the HIV maps, but plotting TB SIR categories
(`adj_TB_YYYY`). Using the same palette and breakpoints as Figure 1 makes
direct visual comparison straightforward — any county that appears red in both
figures has co-elevated HIV and TB burden, which directly motivates the use of
a joint disease model.

```{r tb-sir-maps, fig.cap="Figure 2. TB Standardised Incidence Ratio (SIR) by county, 2012-2017. Same colour scale as Figure 1 to allow direct comparison with HIV burden patterns."}
tb_panels <- lapply(years, function(y) {
  spplot(kenya.one,
         zcol         = paste0("adj_TB_", y),
         col.regions  = my.palette.yor,
         colorkey     = list(space = "right", height = 1.0),
         cuts         = 2,
         main         = as.character(y),
         par.settings = list(fontsize = list(text = 12)))
})

do.call(grid.arrange, c(tb_panels, list(nrow = 2, ncol = 3)))
```

**Output:** A 2 × 3 panel figure of six TB SIR maps. Because TB spreads through
respiratory transmission and has different risk drivers than HIV, the geographic
pattern may differ in some counties. However, spatial overlap between the two
figures — counties appearing red in both — indicates the shared co-epidemic
burden that the INLA joint model in the next section is designed to quantify
and separate into shared versus disease-specific components.

---

# INLA Model: Bayesian Joint Spatio-Temporal Analysis

The descriptive SIR maps above do not account for spatial autocorrelation
(neighbouring counties tend to have similar rates), temporal autocorrelation
(smooth year-to-year trends), or the correlation *between* HIV and TB. The
Bayesian joint model below addresses all three simultaneously using R-INLA.

The model is a **shared-component spatio-temporal Poisson model**:

$$
O_{ist} \mid \mu_{ist} \sim \text{Poisson}(\mu_{ist}), \qquad
\log(\mu_{ist}) = \log(E_{ist}) + \alpha_s + u_{is} + \beta_s^{sp} \cdot \phi_i
+ v_{is} + \beta_s^{tm} \cdot \gamma_t + \epsilon_{ist} + \beta_s^{ST} \cdot \psi_{it}
$$

where $\alpha_s$ is a disease-specific intercept, $u_{is}$ and $v_{is}$ are
disease-specific and shared spatial random effects, $\gamma_t$ and $\psi_{it}$
are shared temporal and space-time fields, and $\beta_s$ are scaling
coefficients (estimated from the data) that quantify how strongly each disease
relates to the shared components.

---

## Spatial Adjacency Graph

**What this chunk does:**
The Besag spatial random effect in INLA requires knowing which counties share a
boundary (i.e. are spatial neighbours). `poly2nb()` reads the county polygons
and builds a neighbourhood list using **queen contiguity** — two counties are
neighbours if they share any boundary point or vertex (not just a full edge).
`nb2INLA()` writes this list to a text file in INLA's specific graph format.
The path to this file is stored in `kenya.adj` for later reference. The average
number of neighbours tells us how connected the spatial graph is — counties
with more neighbours have their estimates smoothed more strongly toward the
regional average.

```{r adjacency-graph}
# Build the queen-contiguity neighbourhood list from county boundary polygons
adj_graph <- poly2nb(kenya.one)

# Write the neighbourhood structure to an INLA-format graph text file
nb2INLA("kenya.graph", adj_graph)
kenya.adj <- file.path(getwd(), "kenya.graph")

cat("Total counties in adjacency graph:", length(adj_graph), "\n")
cat("Average number of neighbours:     ",
    round(mean(sapply(adj_graph, length)), 2), "\n")
```

**Output:** Two lines, e.g.:
```
Total counties in adjacency graph: 47
Average number of neighbours:      5.49
```
A county with many neighbours (e.g. a central county like Nakuru) will have its
spatial estimate smoothed more toward the regional average than a border or
island county with few neighbours (e.g. Mombasa). The file `kenya.graph` is
also written to the working directory.

---

## Reshape Data to Long Format

**What this chunk does:**
INLA requires the data in **long format**: one row per county × year × disease
combination. With 47 counties, 6 years, and 2 diseases, the long-format data
frame has 47 × 6 × 2 = **564 rows**. Reshaping proceeds in two steps:
(1) the wide-format `data_pop` is unstacked into county-year vectors using
`as.vector(as.matrix())`, which reads the matrix column by column (i.e. year
by year within each county); (2) HIV and TB rows are stacked vertically using
`c()`. The `Disease` column labels which rows belong to which disease, and
`Expected` and `Observed` are the Poisson offset and response respectively.

```{r reshape-long}
n_counties <- nrow(data_pop)   # 47

# ── Step 1: Unstack wide data into year-ordered vectors (reads column by column)
observed.HIV   <- as.vector(as.matrix(data_pop[, paste0("obs.HIV", years)]))
observed.TB    <- as.vector(as.matrix(data_pop[, paste0("obs.TB",  years)]))
expected.HIV.v <- as.vector(as.matrix(data_pop[, paste0("exp.HIV", years)]))
expected.TB.v  <- as.vector(as.matrix(data_pop[, paste0("exp.TB",  years)]))
adj.HIV.rate   <- as.vector(as.matrix(data_pop[, paste0("adj.HIV.", years)]))
adj.TB.rate    <- as.vector(as.matrix(data_pop[, paste0("adj.TB.",  years)]))

Year.vec   <- rep(years, each = n_counties)              # 282 elements
County.vec <- as.factor(rep(data_pop[, 1], length(years)))

# ── Step 2: County x year data frame (282 rows = 47 counties x 6 years) ──────
Joint_data <- data.frame(
  County.ID = as.numeric(County.vec),
  Year.ID   = Year.vec,
  O.HIV     = observed.HIV,
  O.TB      = observed.TB,
  E.HIV     = expected.HIV.v,
  E.TB      = expected.TB.v,
  SIR.HIV   = adj.HIV.rate,
  SIR.TB    = adj.TB.rate
)

# ── Step 3: Stack HIV rows (1-282) and TB rows (283-564) ─────────────────────
n.st <- nrow(Joint_data)   # 282 county-year units per disease

data <- data.frame(
  County.ID      = rep(Joint_data$County.ID, 2),
  Year           = rep(Joint_data$Year.ID,   2),
  Year.ID        = rep(Joint_data$Year.ID,   2),
  County.Year.ID = rep(seq_len(n.st),        2)
)

data$Expected <- with(Joint_data, c(E.HIV, E.TB))  # First 282 = HIV; next 282 = TB
data$Observed <- with(Joint_data, c(O.HIV, O.TB))
data$Disease  <- rep(c("HIV", "TB"), each = n.st)

cat("Long-format data frame:", nrow(data), "rows x", ncol(data), "columns\n")
cat("HIV rows:", sum(data$Disease == "HIV"),
    "| TB rows:", sum(data$Disease == "TB"), "\n")
```

**Output:** Two confirmation lines, e.g.:
```
Long-format data frame: 564 rows x 8 columns
HIV rows: 282 | TB rows: 282
```
The equal split confirms HIV and TB are symmetrically represented. The first
282 rows are HIV (counties 1–47 cycling through years 2012–2017); the next 282
are TB in the same order. This structure is required for the INLA joint model.

---

## Index Construction

**What this chunk does:**
INLA identifies each random effect through an integer index column in the data
frame. The key design principle is the **NA structure**: for a disease-specific
effect, the index is active (positive integer) only for that disease's rows,
and is `NA` for all other rows. INLA treats NA entries as observations that do
not contribute to that particular random effect. Shared effects use a single
index active for both diseases. The `ifelse()` calls implement this NA-masking
efficiently. All index columns created here are passed to the model formula
in the next section.

```{r build-indices}
# ── Spatial index: county code 1-47, repeated across years and diseases ───────
data$sp.idx <- rep(1:n_counties, length(years) * 2)

# ── Temporal index: maps calendar year to integer 1-6 ────────────────────────
# 2012 → 1, 2013 → 2, ..., 2017 → 6
data$tm.idx <- data$Year - 2011

# ── Disease index and intercept factor ────────────────────────────────────────
data$dis.idx   <- rep(1:2, each = n.st)          # 1 = HIV, 2 = TB
data$intercept <- as.factor(data$dis.idx)         # Disease-specific intercepts

# ── Space-time interaction index: unique ID for each county x year cell ───────
data$sp.tm.idx <- rep(seq_len(n.st), 2)

# ── Shared effect indices (NA structure: INLA uses these for both diseases) ───
data$sp.dum    <- NA                              # Shared spatial field (NA = copy mechanism handles it)
data$tm.dum    <- NA                              # Shared temporal field
data$sp.tm.dum <- data$County.Year.ID            # Shared space-time field

# ── HIV-specific spatial index (NA for all TB rows) ──────────────────────────
data$sp.1    <- ifelse(data$dis.idx == 1, data$sp.idx, NA)
data$sp.idx1 <- data$sp.1    # Alias used in the formula

# ── TB-specific spatial index (NA for all HIV rows) ──────────────────────────
data$sp.2    <- ifelse(data$dis.idx == 2, data$sp.idx, NA)
data$sp.idx2 <- data$sp.2

# ── HIV-specific temporal index ───────────────────────────────────────────────
data$tm.1    <- ifelse(data$dis.idx == 1, data$tm.idx, NA)
data$tm.idx1 <- data$tm.1

# ── TB-specific temporal index ────────────────────────────────────────────────
data$tm.2    <- ifelse(data$dis.idx == 2, data$tm.idx, NA)
data$tm.idx2 <- data$tm.2

# ── HIV-specific space-time index ─────────────────────────────────────────────
data$sp.tm.1    <- ifelse(data$dis.idx == 1, data$sp.tm.idx, NA)
data$sp.tm.idx1 <- data$sp.tm.1

# ── TB-specific space-time index ──────────────────────────────────────────────
data$sp.tm.2    <- ifelse(data$dis.idx == 2, data$sp.tm.idx, NA)
data$sp.tm.idx2 <- data$sp.tm.2

# Preview the index structure
cat("Index structure (first 6 rows):\n")
print(head(data[, c("County.ID","Year","Disease","dis.idx",
                    "sp.idx","tm.idx","sp.tm.idx")]))
```

**Output:** A 6-row table showing the key index columns for the first six
observations (all HIV, county 1–6, year 2012):
```
  County.ID Year Disease dis.idx sp.idx tm.idx sp.tm.idx
1         1 2012     HIV       1      1      1         1
2         2 2012     HIV       1      2      1         2
...
```
Confirming that `sp.idx` cycles through 1–47 per year, `tm.idx = 1` for 2012,
and `Disease = "HIV"` for the first 282 rows is essential before proceeding.
An error in the index construction would silently mis-assign random effects to
the wrong counties or time points.

---

## Adjacency Matrices

**What this chunk does:**
The INLA random effects require the spatial and temporal dependency structures
expressed as matrices. The **spatial adjacency matrix** (`w.sp`) is a binary
47 × 47 matrix where entry (i,j) = 1 if counties i and j share a boundary,
and 0 otherwise. The **temporal adjacency matrix** (`w.tm`) is a 6 × 6 matrix
encoding a first-order random walk: each year connects only to the year
immediately before and after it, imposing a smooth temporal trend. The boundary
years (2012, 2017) each connect to only one neighbour.

```{r adjacency-matrices}
# ── Spatial adjacency: binary 47x47 matrix from the neighbourhood list ────────
# style = "B" = binary (0/1 entries); zero.policy = TRUE handles any county
# with no neighbours without throwing an error
w.sp <- as(nb2mat(adj_graph, style = "B", zero.policy = TRUE), "Matrix")

# ── Temporal adjacency: first-order random walk structure (6x6) ───────────────
n_years <- length(years)                       # 6
w.tm    <- Diagonal(n_years, x = 0)           # Start with a zero diagonal matrix

# Connect each year to its immediate neighbours only
w.tm[1, 2]                 <- 1               # 2012 connects to 2013
w.tm[n_years, n_years - 1] <- 1               # 2017 connects to 2016

for (i in 2:(n_years - 1)) {
  w.tm[i, i - 1] <- 1                          # Year t connects to t-1
  w.tm[i, i + 1] <- 1                          # Year t connects to t+1
}

cat("Spatial adjacency matrix:  ", nrow(w.sp), "x", ncol(w.sp), "\n")
cat("Temporal adjacency matrix: ", nrow(w.tm), "x", ncol(w.tm), "\n")
```

**Output:**
```
Spatial adjacency matrix:   47 x 47
Temporal adjacency matrix:   6 x 6
```
These matrices are passed to the `graph =` argument in the INLA formula. The
spatial matrix tells the model *which* counties are spatially correlated; the
temporal matrix tells it *which* years are temporally correlated. Together they
define the dependency structure that enables the model to borrow strength across
space and time.

---

## Prior Specifications

**What this chunk does:**
Prior distributions encode our uncertainty about model parameters before
seeing the data. Two types of prior are specified. The **log-Gamma prior on
precision** (inverse variance) is placed on all random effect variance
hyperparameters. With shape = 0.5 and rate = 0.0005, this is weakly
informative — it allows large variances and does not strongly constrain the
estimates. The **Normal prior on the copy-effect beta** is placed on the
scaling coefficients linking each disease to the shared components. A
precision of 1/5.9 gives a standard deviation of ~2.43, which is very wide
and allows any plausible scaling value. The range constraint `c(0, Inf)` forces
beta > 0, meaning HIV and TB must have *positive* associations with the shared
spatial/temporal field.

```{r priors}
# ── Log-Gamma prior on precision hyperparameters (all random effects) ─────────
# Shape = 0.5, rate = 0.0005 → weakly informative; allows large variances
prior.prec <- list(prior   = "loggamma",
                   param   = c(0.5, 0.0005),
                   initial = 0)

# ── Normal prior on copy-effect scaling betas ─────────────────────────────────
# Mean = 0, precision = 1/5.9 (sd ≈ 2.43) — wide, weakly informative
# Range c(0, Inf): both diseases must positively load onto the shared component
prior.beta.s   <- list(prior   = "normal", param = c(0, 1/5.9),
                       fixed   = FALSE, initial = 0.01)   # Spatial copy beta

prior.beta.t   <- list(prior   = "normal", param = c(0, 1/5.9),
                       fixed   = FALSE, initial = 0.01)   # Temporal copy beta

prior.beta.s.t <- list(prior   = "normal", param = c(0, 1/5.9),
                       fixed   = FALSE, initial = 0.01)   # Space-time copy beta

# Model scaling: FALSE = do not rescale random effects to unit variance
inla.scale <- FALSE
```

**Output:** No visible output. Five prior specification lists stored in memory.
Sensitivity analyses can be conducted by modifying the `param` values here and
re-running the model. The `initial` values provide starting points for the
optimisation within INLA and generally do not affect the final posterior.

---

## Model Formula

**What this chunk does:**
The INLA model formula specifies the full structure of the linear predictor.
`-1` suppresses the global intercept; `intercept` (a two-level factor) provides
disease-specific intercepts instead. Each `f()` term defines one random effect:
`model = "besag"` specifies the intrinsic CAR (Conditionally AutoRegressive)
model for spatially or temporally structured effects (using the adjacency
matrices as the graph); `model = "iid"` specifies an unstructured
(independent) random effect for the space-time interaction. The `copy`
mechanism is the defining feature of this joint model: `f(sp.1, copy = "sp.dum")`
tells INLA that the HIV spatial index (`sp.1`) is a *scaled copy* of the shared
spatial field (`sp.dum`), with the scaling factor beta estimated from the data.
A large beta means HIV strongly follows the shared spatial pattern; a small
beta means HIV has mostly disease-specific spatial variation.

```{r model-formula}
formula <- Observed ~ -1 + intercept +

  # ── Disease-specific spatial random effects (Besag intrinsic CAR) ─────────
  f(sp.idx1, model = "besag", graph = w.sp,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +   # HIV only
  f(sp.idx2, model = "besag", graph = w.sp,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +   # TB only

  # ── Shared spatial field + scaled copies for HIV and TB ───────────────────
  f(sp.dum,  model = "besag", graph = w.sp,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +   # Shared field
  f(sp.1, copy = "sp.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s)) +                            # HIV scaling
  f(sp.2, copy = "sp.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s)) +                            # TB scaling

  # ── Disease-specific temporal effects (Besag on first-order RW graph) ─────
  f(tm.idx1, model = "besag", graph = w.tm,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +   # HIV only
  f(tm.idx2, model = "besag", graph = w.tm,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +   # TB only

  # ── Shared temporal field + scaled copies ─────────────────────────────────
  f(tm.dum,  model = "besag", graph = w.tm,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +   # Shared field
  f(tm.1, copy = "tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.t)) +                            # HIV scaling
  f(tm.2, copy = "tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.t)) +                            # TB scaling

  # ── Disease-specific space-time IID interactions ──────────────────────────
  f(sp.tm.idx1, model = "iid", hyper = list(prec = prior.prec)) +  # HIV only
  f(sp.tm.idx2, model = "iid", hyper = list(prec = prior.prec)) +  # TB only

  # ── Shared space-time IID field + scaled copies ───────────────────────────
  f(sp.tm.dum,  model = "iid", hyper = list(prec = prior.prec)) +  # Shared
  f(sp.tm.1, copy = "sp.tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s.t)) +                          # HIV scaling
  f(sp.tm.2, copy = "sp.tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s.t))                            # TB scaling
```

**Output:** No visible output. The formula object is stored in memory. In total
the formula specifies 15 random effect terms: three disease-specific effects
(spatial, temporal, space-time) × 2 diseases, plus three shared fields, plus
three pairs of copy terms. This is the most complex part of the model — each
term is labelled with comments to make the structure transparent.

---

## Model Fitting

**What this chunk does:**
`inla()` fits the Bayesian model using Integrated Nested Laplace Approximation,
which is far faster than MCMC for models of this type. Key arguments: `family =
"poisson"` specifies the Poisson response distribution; `E = Expected` provides
the expected counts as a multiplicative offset (so the model estimates the log
relative risk, not the raw log mean); `control.compute` requests DIC (Deviance
Information Criterion, for model comparison), CPO (Conditional Predictive
Ordinate, for out-of-sample fit), and the full posterior configuration for later
marginal calculations. `control.inla` sets the Laplace approximation strategy
with 21 integration points, providing higher accuracy than the default. The
`cache = TRUE` option in the chunk header means this chunk's result is stored
on disk — if you re-knit without changing this chunk, R loads the cached result
instead of re-running INLA (important given the ~20–40 minute run time). The
fitted model is saved to disk separately as an `.RData` file.

```{r fit-model, cache=TRUE}
model.inla <- inla(
  formula,
  family            = "poisson",
  data              = data,
  E                 = Expected,              # Expected counts as Poisson offset
  control.predictor = list(compute = TRUE),  # Compute fitted values
  verbose           = FALSE,                 # Set TRUE to monitor INLA progress
  control.compute   = list(config = TRUE,    # Full posterior config for marginals
                           dic    = TRUE,    # Deviance Information Criterion
                           cpo    = TRUE),   # Conditional Predictive Ordinate
  control.inla      = list(strategy = "laplace", npoints = 21)  # High-accuracy approximation
)

# Save fitted model for reuse in future sessions (avoids re-running INLA)
save(model.inla, file = "st_model_GAMMA.RData")

# Print the model summary
summary(model.inla)
```

**Output:** The INLA model summary table, which contains:

- **Fixed effects**: the two disease-specific intercepts (on the log scale).
  These represent the baseline log relative risk for each disease after all
  random effects are accounted for.
- **Model hyperparameters**: posterior means and 95% credible intervals for
  the precision of each spatial, temporal, and space-time random effect, and
  for the copy-effect betas ($\beta_s$). A large precision means small variance
  (tightly constrained spatial/temporal pattern); a small precision means large
  variance (highly variable pattern). A beta close to 1.0 means the disease
  closely follows the shared component; beta close to 0 means little shared
  pattern.
- **DIC**: lower DIC indicates better model fit.
- **Marginal log-likelihood**: used for formal model comparison.

---

# Results: Spatial Effects

## Extract Posterior Mean Spatial Relative Risks

**What this chunk does:**
The INLA model stores posterior summaries for every random effect in
`model.inla$summary.random`. We extract the posterior **mean** (the single
best-estimate value) for each spatial effect and exponentiate it (because the
model is on the log scale) to obtain the **Relative Risk (RR)** on the
natural scale. Three types of spatial RR are computed: (1) the **shared
spatial RR** common to both diseases; (2) the **disease-specific spatial RRs**,
which capture how each disease departs from the shared pattern; and (3) the
**combined total spatial RR**, which is the sum (on the log scale, i.e. product
on the RR scale) of the disease-specific and the scaled shared component,
representing the full county-level spatial effect for each disease.

```{r extract-spatial}
# Shared spatial RR: exp(posterior mean of the shared spatial field)
kenya.one$SHARED_C <- exp(model.inla$summary.random$sp.dum[,  "mean"])

# Disease-specific spatial RRs (residual spatial pattern after removing shared)
kenya.one$HIV_C    <- exp(model.inla$summary.random$sp.idx1[, "mean"])
kenya.one$TB_C     <- exp(model.inla$summary.random$sp.idx2[, "mean"])

# Combined total spatial RR = disease-specific + scaled shared copy
# On the log scale these add; on the RR scale they multiply
kenya.one$Total.HIV_C <- exp(model.inla$summary.random$sp.idx1[, "mean"] +
                               model.inla$summary.random$sp.1[,   "mean"])

kenya.one$Total.TB_C  <- exp(model.inla$summary.random$sp.idx2[, "mean"] +
                               model.inla$summary.random$sp.2[,   "mean"])
```

**Output:** No visible output. Five numeric columns are added to `kenya.one`
(one value per county). All values are on the relative risk scale: > 1 means
the spatial effect elevates the county's expected count above the national
average; < 1 means the spatial effect is "protective" (below average). These
are the values that will be mapped in the following chunks.

---

## Classify Spatial RRs into Categories

**What this chunk does:**
The continuous posterior mean spatial RRs are cut into four ordered categories
using the same `cut()` approach as the descriptive SIR maps. Note that the
third breakpoint here is 1.00 (not 0.90 as in the SIR maps), reflecting that
model-based RRs are centred around 1.0 and the critical distinction is whether
a county's spatial effect is above or below the national average. The reusable
`classify_rr()` function keeps the code concise.

```{r classify-spatial}
# Breakpoints for model-based posterior spatial RR maps
spatial.breaks <- c(0, 0.40, 0.70, 1.00, 6)
spatial.labels <- c("0.00-0.40", "0.41-0.70", "0.71-1.00", ">1.00")

classify_rr <- function(x) {
  cut(x, breaks = spatial.breaks, include.lowest = TRUE, labels = spatial.labels)
}

kenya.one$SHARED    <- classify_rr(kenya.one$SHARED_C)
kenya.one$HIV       <- classify_rr(kenya.one$HIV_C)
kenya.one$TB        <- classify_rr(kenya.one$TB_C)
kenya.one$Total.HIV <- classify_rr(kenya.one$Total.HIV_C)
kenya.one$Total.TB  <- classify_rr(kenya.one$Total.TB_C)
```

**Output:** No visible output. Five categorical (factor) columns are added to
`kenya.one`. The ">1.00" category marks counties where the posterior spatial RR
exceeds 1 — these are the counties flagged as having above-average spatial risk
after Bayesian smoothing.

---

## Disease-Specific and Shared Spatial Effects

**What this chunk does:**
This three-panel figure decomposes the overall spatial pattern into its shared
and disease-specific components. The **Shared** panel shows counties where
both HIV and TB are jointly elevated or suppressed — these counties have a
common underlying risk driver (e.g. poverty, healthcare access, mobility
patterns). The **HIV** and **TB** panels show the residual disease-specific
spatial risk *after* removing the shared component — these represent risk
factors unique to each disease (e.g. HIV risk from sexual behaviour patterns;
TB risk from indoor crowding and ventilation).

```{r spatial-effects-map, fig.cap="Figure 3. Posterior mean model-based spatial relative risks separated into three components: Shared (left), HIV-specific residual (centre), TB-specific residual (right). Red counties have posterior RR > 1.0; yellow counties have RR < 0.40."}
grid.arrange(
  spplot(kenya.one, zcol = "SHARED", col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 2,
         main = "Shared", par.settings = list(fontsize = list(text = 12))),
  spplot(kenya.one, zcol = "HIV",    col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 2,
         main = "HIV",   par.settings = list(fontsize = list(text = 12))),
  spplot(kenya.one, zcol = "TB",     col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 2,
         main = "TB",    par.settings = list(fontsize = list(text = 12))),
  ncol = 3
)
```

**Output:** A three-panel figure. A county appearing red in the **Shared** panel
has elevated risk for both diseases simultaneously — this is the co-epidemic
signal. A county red only in the **HIV** panel has HIV-specific excess risk
not driven by the shared factor (e.g. a county with high HIV but not
exceptionally high TB). Similarly, counties red only in **TB** represent
TB-specific spatial risk. Comparing the three panels is the main analytical
result of the spatial component of the joint model.

---

## Combined Total Spatial Patterns

**What this chunk does:**
The total spatial effect for each disease integrates both the disease-specific
component and the scaled shared component into a single overall county-level
spatial relative risk. This is the most complete spatial summary of the model:
it shows the full county-level spatial burden for each disease after Bayesian
smoothing, accounting for both what is unique to each disease and what they
share.

```{r total-spatial-map, fig.cap="Figure 4. Combined posterior spatial relative risks for HIV (left) and TB (right): sum of disease-specific and scaled shared spatial components. These represent the full Bayesian-smoothed spatial burden for each disease."}
grid.arrange(
  spplot(kenya.one, zcol = "Total.HIV", col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 6,
         main     = "HIV: Combined Spatial Pattern",
         par.settings = list(fontsize = list(text = 12))),
  spplot(kenya.one, zcol = "Total.TB",  col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 5,
         main     = "TB: Combined Spatial Pattern",
         par.settings = list(fontsize = list(text = 12))),
  ncol = 2
)
```

**Output:** A two-panel figure. The Bayesian smoothing produces cleaner, less
noisy maps than the raw SIR maps in Figures 1–2, because information is
borrowed from spatial neighbours. Small-population counties with erratic raw
SIRs will be pulled toward the regional average here. Counties that remain red
after smoothing are genuinely high-burden areas with strong statistical
evidence of excess risk. The overlap between the HIV and TB maps quantifies
the geographic co-localisation of the two epidemics.

---

# Results: Temporal Trends

Temporal trends show how disease risk changed from 2012 to 2017 at the national
level, after controlling for spatial variation. All trends are on the **Relative
Risk scale** with 95% credible intervals shown as dashed lines. RR = 1 (flat
line) indicates no temporal change in national risk.

**What this chunk does:**
A helper function `get_temporal()` extracts and exponentiates the posterior
mean and 95% credible interval bounds for any temporal random effect from the
INLA model. A second helper `plot_trend()` draws a single time-series panel.
Four panels are produced: (a) the shared trend; (b) the HIV-specific trend;
(c) the TB-specific trend; (d) the combined total trends for HIV and TB
overlaid for direct comparison. The combined trend is computed by summing the
disease-specific temporal effect and the scaled shared temporal copy on the log
scale (equivalent to multiplying the two RRs).

```{r temporal-trends, fig.cap="Figure 5. Posterior mean temporal relative risks 2012-2017 with 95% credible intervals (dashed lines). (a) Shared trend common to both diseases. (b) HIV-specific trend. (c) TB-specific trend. (d) Combined total trends: HIV (red) and TB (blue) overlaid.", fig.height=8}
# Helper: extract and exponentiate posterior mean and 95% CrI for a temporal effect
get_temporal <- function(effect_name) {
  sm <- model.inla$summary.random[[effect_name]]
  list(mean  = exp(sm[, "mean"]),
       lower = exp(sm[, "0.025quant"]),    # Lower bound of 95% credible interval
       upper = exp(sm[, "0.975quant"]))    # Upper bound of 95% credible interval
}

# Extract shared, HIV-specific, and TB-specific temporal effects
shared_t <- get_temporal("tm.dum")    # Shared temporal field
hiv_t    <- get_temporal("tm.idx1")  # HIV-specific temporal field
tb_t     <- get_temporal("tm.idx2")  # TB-specific temporal field

# Combined trends: disease-specific + scaled copy of shared temporal field
hiv_copy <- model.inla$summary.random$tm.1
tb_copy  <- model.inla$summary.random$tm.2

hiv_comb <- list(
  mean  = exp(model.inla$summary.random$tm.idx1[, "mean"]       + hiv_copy[, "mean"]),
  lower = exp(model.inla$summary.random$tm.idx1[, "0.025quant"] + hiv_copy[, "0.025quant"]),
  upper = exp(model.inla$summary.random$tm.idx1[, "0.975quant"] + hiv_copy[, "0.975quant"])
)

tb_comb <- list(
  mean  = exp(model.inla$summary.random$tm.idx2[, "mean"]       + tb_copy[, "mean"]),
  lower = exp(model.inla$summary.random$tm.idx2[, "0.025quant"] + tb_copy[, "0.025quant"]),
  upper = exp(model.inla$summary.random$tm.idx2[, "0.975quant"] + tb_copy[, "0.975quant"])
)

# Helper: draw one temporal trend panel with its 95% credible interval
plot_trend <- function(y_mean, y_lower, y_upper, ylim,
                       main_title, panel_label, line_col = "black") {
  plot(years, y_mean, ylim = ylim, type = "l",
       main = main_title, xlab = panel_label, ylab = "",
       frame = FALSE, col = line_col, lty = 1, lwd = 2, cex.lab = 1.2)
  title(ylab = "Relative Risk", line = 1.9, cex.lab = 1.2)
  lines(years, y_lower, col = line_col, lty = 3, lwd = 0.5)  # Lower 95% CrI
  lines(years, y_upper, col = line_col, lty = 3, lwd = 0.5)  # Upper 95% CrI
}

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# (a) Shared temporal trend (black line; both diseases)
plot_trend(shared_t$mean, shared_t$lower, shared_t$upper,
           ylim = c(0.0, 2.0),
           main_title = "Shared Temporal Trend", panel_label = "(a)")

# (b) HIV-specific temporal trend (red line; departures from shared trend)
plot_trend(hiv_t$mean, hiv_t$lower, hiv_t$upper,
           ylim = c(0.8, 1.2),
           main_title = "HIV Temporal Trend", panel_label = "(b)",
           line_col = "red")

# (c) TB-specific temporal trend (blue line; departures from shared trend)
plot_trend(tb_t$mean, tb_t$lower, tb_t$upper,
           ylim = c(0.8, 1.2),
           main_title = "TB Temporal Trend", panel_label = "(c)",
           line_col = "blue")

# (d) Combined total temporal trends: HIV (red) and TB (blue) overlaid
plot_trend(hiv_comb$mean, hiv_comb$lower, hiv_comb$upper,
           ylim = c(0.5, 1.6),
           main_title = "Combined Temporal Trend", panel_label = "(d)",
           line_col = "red")
lines(years, tb_comb$mean,  col = "blue", lty = 1, lwd = 2)
lines(years, tb_comb$lower, col = "blue", lty = 3, lwd = 0.5)
lines(years, tb_comb$upper, col = "blue", lty = 3, lwd = 0.5)
legend(2012, 1.6, legend = c("HIV", "TB"),
       col = c("red", "blue"), lwd = 2, lty = 1, cex = 0.8)

par(mfrow = c(1, 1))   # Reset graphics to single-panel layout
```

**Output:** A 2 × 2 panel figure with four time-series plots.

- **Panel (a) — Shared trend:** shows whether the risk common to both diseases
  is rising (RR increasing), falling (RR decreasing), or stable over time.
  Wide credible intervals (dashed lines far from the mean) indicate high
  uncertainty about the temporal trend direction.
- **Panels (b) and (c) — Disease-specific trends:** show whether HIV and TB
  diverge from the shared trend. For example, TB declining faster than the
  shared trend would appear as a downward curve in panel (c) even if the shared
  trend is flat.
- **Panel (d) — Combined total trends:** the most policy-relevant output; shows
  the overall national temporal trajectory for each disease on the same scale.
  If HIV (red) is above TB (blue) throughout, HIV carries a consistently higher
  national relative risk. Overlapping credible intervals in the final years
  would suggest the two diseases are converging in their national temporal
  burden.

---

# Results: Space-Time Interaction

The shared space-time interaction captures residual county-year heterogeneity
that is *not* explained by the main spatial pattern or the main temporal trend.
In other words, it identifies county-year combinations that stand out as unusual
even after accounting for the overall spatial and temporal structure — for
example, a county that experiences a sudden unexplained surge in one particular
year.

## Extract Posterior Means of the Space-Time Interaction

**What this chunk does:**
The posterior means of the shared space-time random effect (`sp.tm.dum`) are
extracted from the INLA model output. These values are on the **log scale**
(before exponentiation). They are reshaped into a 47 × 6 matrix so that the
space-time pattern can be examined county by county and year by year. Positive
values indicate a county-year combination with *more* shared disease risk than
expected from the main spatial and temporal effects; negative values indicate
*less*. A preview of the first five counties is printed for verification.

```{r space-time-posterior}
# Extract posterior means of the shared space-time interaction (log scale)
delta_s_vec <- model.inla$summary.random$sp.tm.dum[, "mean"]

# Reshape into a 47-county x 6-year matrix (filled column by column = year by year)
delta.matrix_s           <- matrix(delta_s_vec, nrow = n_counties,
                                   ncol = length(years), byrow = FALSE)
rownames(delta.matrix_s) <- data.kenya$COUNTY
colnames(delta.matrix_s) <- years

cat("Shared space-time interaction matrix (posterior means, log scale):\n")
print(round(delta.matrix_s[1:5, ], 4))
```

**Output:** A 5 × 6 table (first five counties × six years) of log-scale
posterior mean space-time interaction values. Values close to zero indicate
county-year combinations that fit the overall spatial and temporal model well.
A value of +0.3 in a given cell would correspond to an ~35% excess RR
(exp(0.3) ≈ 1.35) beyond what the main spatial and temporal effects predict.
This matrix is the raw input for the exceedance probability maps below.

---

## Exceedance Probabilities P(δ > 0 | data)

**What this chunk does:**
Rather than relying solely on the posterior mean (which does not quantify
uncertainty), the **exceedance probability** provides a formal Bayesian
measure of evidence for excess risk. For each county-year cell,
`inla.pmarginal(0, marg)` computes the posterior probability that the
space-time interaction ≤ 0 (no excess); subtracting from 1 gives P(δ > 0 |
data) — the probability of excess risk. The 282 probabilities are arranged into
a 47 × 6 matrix and categorised into four bands: 0.76–1.00 indicates very
strong evidence of excess risk; 0.00–0.25 indicates very strong evidence of
below-expected risk.

```{r exceedance-prob}
# δ = 0 on the log scale corresponds to RR = 1 (no excess risk)
threshold <- 0

# Compute P(delta > 0 | data) for all 282 county-year cells
# marginals.random[[9]] = the 9th f() term in the formula = sp.tm.dum
pp.delta_s <- sapply(
  model.inla$marginals.random[[9]],
  function(marg) 1 - inla.pmarginal(threshold, marg)
)

# Reshape into county x year matrix
pp.delta.matrix_s <- matrix(pp.delta_s, nrow = n_counties,
                             ncol = length(years), byrow = FALSE)

# Categorise into four probability bands for choropleth mapping
pp.breaks <- c(0, 0.25, 0.50, 0.75, Inf)
pp.labels <- c("0.00-0.25", "0.26-0.50", "0.51-0.75", "0.76-1.00")

# Build a per-year categorical data frame (one column per year)
pp.delta.factor_s <- data.frame(County = data.kenya$COUNTY)

for (i in seq_along(years)) {
  pp.delta.factor_s[[paste0("X2", i + 11)]] <-
    cut(pp.delta.matrix_s[, i],
        breaks = pp.breaks, include.lowest = TRUE, labels = pp.labels)
}

cat("Exceedance probability data frame:", nrow(pp.delta.factor_s),
    "rows x", ncol(pp.delta.factor_s), "columns\n")
```

**Output:** A dimension confirmation, e.g.:
`Exceedance probability data frame: 47 rows x 7 columns` (County + six year
columns). Each cell contains one of four probability band labels. Counties in
the "0.76–1.00" band in a given year have very strong posterior evidence of
unusual excess shared disease risk in that specific year — these are the
highest-priority county-year combinations for public health investigation.

---

## Space-Time Exceedance Probability Maps

**What this chunk does:**
The exceedance probability categories are attached to the Kenya spatial object
and displayed as a 2 × 3 grid of choropleth maps using a Yellow-Green (YlGn)
palette. A different palette (green rather than red) is deliberately used here
to visually distinguish the *exceedance probability maps* (which show
statistical evidence) from the *SIR and RR maps* earlier (which show magnitude
of risk). Darker green counties in a given year have strong posterior evidence
of unusual excess shared disease burden in that county and year — beyond what
the general spatial and temporal model can explain.

```{r space-time-maps, fig.cap="Figure 6. Posterior probability P(shared space-time interaction > 0 | data) by county, 2012-2017. Dark green = probability > 0.75 (strong evidence of excess county-year shared risk). Yellow-green = probability < 0.25 (evidence of below-expected risk)."}
# Attach the exceedance probability columns to the kenya spatial object
attr(kenya, "data") <- data.frame(data.kenya, pp.delta.factor_s)

# Yellow-Green palette: low exceedance (yellow) to high exceedance (dark green)
my.palette.ygn <- brewer.pal(4, "YlGn")

# Build one map panel per year and arrange in a 2x3 grid
pp_panels <- lapply(seq_along(years), function(i) {
  zcol <- paste0("X2", i + 11)    # Column names: X212, X213, ..., X217
  spplot(obj = kenya, zcol = zcol,
         col.regions  = my.palette.ygn, cuts = 2,
         colorkey     = list(space = "right", height = 1.0),
         main         = as.character(years[i]),
         par.settings = list(fontsize = list(text = 12)))
})

do.call(grid.arrange, c(pp_panels, list(nrow = 2, ncol = 3)))
```

**Output:** A 2 × 3 panel figure showing the exceedance probability for each
of the six study years. Counties that are consistently dark green across
multiple years are experiencing residual excess shared risk that cannot be
explained by the overall spatial or temporal patterns — these are the counties
warranting the most urgent public health attention. Counties that are yellow
across all years fit the model's expected pattern well. The year-to-year
variation in which counties appear dark green reveals whether the unusual risk
is persistent or episodic.

---

# Session Information

**What this chunk does:**
`sessionInfo()` records the exact R version, operating system, and package
versions used to produce this document. This is an essential component of a
reproducible research report — if results differ between two runs of the same
code, version differences in packages (particularly INLA, which is updated
frequently) are a common cause. Always include this output when sharing the
document.

```{r session-info}
sessionInfo()
```

**Output:** A detailed printout of the complete R session environment,
including: R version number, platform and operating system, locale settings,
base packages, and all attached packages with their exact version numbers. When
submitting or sharing this analysis, this section allows others to replicate
the exact software environment that produced these results.
