---
title: "Spatio-Temporal Analysis of HIV and Tuberculosis Co-Morbidity in Kenya"
subtitle: "County-Level Bayesian Joint Disease Modelling, 2012–2017"
author: "Verrah Otiende"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
# ── Global chunk options ──────────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo      = TRUE,    # Show code in output
  message   = FALSE,   # Suppress package messages
  warning   = FALSE,   # Suppress warnings
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)
```

---

# Overview

This document presents a Bayesian spatio-temporal **joint disease model** for
HIV and Tuberculosis (TB) across Kenya's **47 counties** from **2012 to 2017**.
The analysis proceeds in five stages:

1. **Data import** — load raw case counts and population projections
2. **Indirect standardisation** — compute Standardised Incidence Ratios (SIRs)
   for descriptive mapping
3. **Spatial data preparation** — merge tabular data with county shapefiles
4. **INLA model fitting** — fit a shared-component spatio-temporal Poisson model
5. **Results extraction and visualisation** — spatial maps, temporal trends, and
   space-time interaction probabilities

---

# Package Dependencies

All packages are loaded here so that dependencies are transparent and the
workspace can be reproduced in a single step.

```{r packages}
# ── Spatial data handling ─────────────────────────────────────────────────────
library(sp)          # Spatial objects (SpatialPolygonsDataFrame)
library(sf)          # Simple features (sf objects)
library(rgdal)       # Read/write spatial data formats, including .shp
library(spdep)       # Spatial dependence: weights, autocorrelation
library(maptools)    # Map tools and spatial utilities
library(spacetime)   # Space-time data structures

# ── Mapping and visualisation ─────────────────────────────────────────────────
library(maps)        # Base map plotting functions
library(mapdata)     # High-resolution county boundary points
library(RColorBrewer)# Colour palettes for choropleth maps
library(lattice)     # Trellis graphics (used by spplot)
library(latticeExtra)# Extensions to lattice (panel functions, etc.)
library(gridExtra)   # Arrange multiple grid graphics on one page
library(grid)        # Low-level grid graphics system
library(car)         # Companion to Applied Regression (misc. plot helpers)

# ── Epidemiological and statistical tools ─────────────────────────────────────
library(epitools)    # Epidemiological tools (rate calculations)
library(DCluster)    # Detection of spatial disease clusters
library(stats)       # Base R statistical functions
library(plotrix)     # Miscellaneous plot utilities

# ── Bayesian spatio-temporal modelling ───────────────────────────────────────
library(INLA)        # Integrated Nested Laplace Approximation

# ── Reporting ─────────────────────────────────────────────────────────────────
library(knitr)       # Dynamic report generation
```

---

# Data Import

## File Paths

Update the paths in this single block to match your local directory structure.
All subsequent code reads from these variables, so only this section needs
to change when moving the project to a new machine.

```{r file-paths}
# ── User-defined paths (edit here only) ──────────────────────────────────────
PATH_HIV <- "data/HIV_Data_2018.csv"
PATH_TB  <- "data/TB_Data_2018.csv"
PATH_SHP <- "data/County_Kenya/County.shp"
```

## Load Case Count Data

```{r load-data}
# Raw HIV new-diagnosis counts and projected populations, 2012–2017
# Rows: 47 counties | Columns: County code, name, annual cases (HIV_YYYY),
#                              annual population (PP_YYYY), totals
HIV_data <- read.csv(PATH_HIV)

# Raw TB notification counts and projected populations, 2012–2017
# Same structure as HIV_data; County_ID plays the role of Sn.no
TB_data  <- read.csv(PATH_TB)

# Quick sanity check: both files should have 47 rows (one per county)
stopifnot(nrow(HIV_data) == 47, nrow(TB_data) == 47)

cat("HIV data: ", nrow(HIV_data), "counties,", ncol(HIV_data), "columns\n")
cat("TB  data: ", nrow(TB_data),  "counties,", ncol(TB_data),  "columns\n")
```

---

# Indirect Standardisation and SIR Computation

Standardised Incidence Ratios (SIRs) are computed using **indirect
standardisation**:

$$
\text{SIR}_{ic} = \frac{O_{ic}}{E_{ic}}, \quad
E_{ic} = \bar{\lambda}_c \times \bar{P}_i
$$

where $O_{ic}$ is the observed count for county $i$ in period $c$,
$\bar{\lambda}_c$ is the national crude rate for period $c$, and
$\bar{P}_i = P_{i,\text{Total}} / 6$ is the county's average annual
population (the *standard population*).

## Standard Populations

```{r standard-populations}
# Average annual projected population across the six study years.
# Used as the reference denominator when computing expected counts.
standard.HIV <- HIV_data$PP_Total / 6
standard.TB  <- TB_data$PP_Total  / 6
```

## National Crude Incidence Rates

Rather than repeating the same expression six times, we compute all annual
crude rates in a single vectorised step and store them in named vectors.

```{r crude-rates}
years <- 2012:2017   # Study period

# --- HIV crude rates ---------------------------------------------------------
# National crude rate = (total national cases) / (total national person-years)
cruderate.HIV <- sapply(years, function(y) {
  col_cases <- paste0("HIV_", y)
  col_pop   <- paste0("PP_",  y)
  sum(HIV_data[[col_cases]]) / sum(HIV_data[[col_pop]])
})
names(cruderate.HIV) <- years

# Pooled rate across all six years
cruderate.HIV_Total <- sum(HIV_data$HIV_Total) / sum(HIV_data$PP_Total)

# --- TB crude rates ----------------------------------------------------------
cruderate.TB <- sapply(years, function(y) {
  col_cases <- paste0("TB_", y)
  col_pop   <- paste0("PP_", y)
  sum(TB_data[[col_cases]]) / sum(TB_data[[col_pop]])
})
names(cruderate.TB) <- years

cruderate.TB_Total <- sum(TB_data$TB_Total) / sum(TB_data$PP_Total)

# Print a summary table
crude_summary <- data.frame(
  Year          = years,
  HIV_CrudeRate = cruderate.HIV,
  TB_CrudeRate  = cruderate.TB
)
knitr::kable(crude_summary, digits = 6,
             caption = "National annual crude incidence rates (cases per person-year)")
```

## Expected Counts

```{r expected-counts}
# Expected cases = national crude rate x county standard population.
# Returns a 47-row matrix (counties x years) for each disease.

# --- HIV expected counts -----------------------------------------------------
expected.HIV <- sapply(years, function(y) {
  cruderate.HIV[as.character(y)] * standard.HIV
})
colnames(expected.HIV) <- years   # columns = years, rows = counties

expected.HIV_Total <- cruderate.HIV_Total * standard.HIV

# --- TB expected counts ------------------------------------------------------
expected.TB <- sapply(years, function(y) {
  cruderate.TB[as.character(y)] * standard.TB
})
colnames(expected.TB) <- years

expected.TB_Total <- cruderate.TB_Total * standard.TB
```

## SIR Computation (Continuous)

```{r sir-continuous}
# SIR_iy = observed_iy / expected_iy  (county i, year y)
# Values > 1 indicate above-national-average burden.

# --- HIV SIRs ----------------------------------------------------------------
adj.HIV <- sapply(years, function(y) {
  HIV_data[[paste0("HIV_", y)]] / expected.HIV[, as.character(y)]
})
colnames(adj.HIV) <- years

adj.HIV.total <- HIV_data$HIV_Total / expected.HIV_Total

# --- TB SIRs -----------------------------------------------------------------
adj.TB <- sapply(years, function(y) {
  TB_data[[paste0("TB_", y)]] / expected.TB[, as.character(y)]
})
colnames(adj.TB) <- years

adj.TB.total <- TB_data$TB_Total / expected.TB_Total
```

## SIR Classification (Categorical)

Continuous SIRs are discretised into four ordered categories for choropleth
mapping. The same breakpoints are applied uniformly across all years and both
diseases.

```{r sir-categorical}
# Breakpoints and labels applied to all year/disease combinations
sir.breaks  <- c(0, 0.40, 0.70, 0.90, 6)
sir.labels  <- c("0.00-0.40", "0.41-0.70", "0.71-0.99", "1.00+")

# Helper: classify a numeric SIR vector into the four categories
classify_sir <- function(x) {
  cut(x, breaks = sir.breaks, include.lowest = TRUE, labels = sir.labels)
}

# --- HIV categorical SIRs ----------------------------------------------------
adj_HIV_cat <- apply(adj.HIV, 2, classify_sir)   # 47 x 6 factor matrix
colnames(adj_HIV_cat) <- paste0("adj_HIV_", years)

adj_HIV_total_cat <- classify_sir(adj.HIV.total)

# --- TB categorical SIRs -----------------------------------------------------
adj_TB_cat <- apply(adj.TB, 2, classify_sir)
colnames(adj_TB_cat) <- paste0("adj_TB_", years)

adj_TB_total_cat <- classify_sir(adj.TB.total)
```

## Build the Consolidated Analysis Dataset

```{r build-data-prop}
# Flatten the matrix columns into individual named vectors so they can be
# passed cleanly to data.frame().  Column names mirror the original script.

# Observed counts
obs_HIV <- setNames(as.data.frame(
  sapply(years, function(y) HIV_data[[paste0("HIV_", y)]])),
  paste0("obs.HIV", years))

obs_TB  <- setNames(as.data.frame(
  sapply(years, function(y) TB_data[[paste0("TB_", y)]])),
  paste0("obs.TB", years))

# Expected counts
exp_HIV <- setNames(as.data.frame(expected.HIV), paste0("exp.HIV", years))
exp_TB  <- setNames(as.data.frame(expected.TB),  paste0("exp.TB",  years))

# Continuous SIRs
sir_HIV <- setNames(as.data.frame(adj.HIV), paste0("adj.HIV.", years))
sir_TB  <- setNames(as.data.frame(adj.TB),  paste0("adj.TB.",  years))

# Categorical SIRs
cat_HIV <- setNames(as.data.frame(adj_HIV_cat), paste0("adj_HIV_", years))
cat_TB  <- setNames(as.data.frame(adj_TB_cat),  paste0("adj_TB_",  years))

# Assemble into one wide data frame (one row per county)
data_prop <- data.frame(
  County.ID    = HIV_data$Sn.no,
  County       = HIV_data$County,
  obs_HIV,
  obs.HIVtotal = HIV_data$HIV_Total,
  obs_TB,
  obs.TBtotal  = TB_data$TB_Total,
  exp_HIV,
  exp.HIVtotal = expected.HIV_Total,
  exp_TB,
  exp.TBtotal  = expected.TB_Total,
  sir_HIV,
  adj.HIV.total,
  sir_TB,
  adj.TB.total,
  cat_HIV,
  adj_HIV_total_cat,
  cat_TB,
  adj_TB_total_cat
)

cat("data_prop dimensions:", nrow(data_prop), "x", ncol(data_prop), "\n")
```

---

# Spatial Data Preparation

## Load and Align Shapefile

```{r load-shapefile}
# Read the Kenya county polygon shapefile.
# The attribute table field "COUNTY" must match the County names in data_prop.
kenya <- readOGR(PATH_SHP, verbose = FALSE)

# Reorder data_prop rows to match the shapefile's county ordering.
# This is essential before the merge so that spatial and attribute data align.
shp_order <- match(kenya$COUNTY, data_prop$County)
data_pop   <- data_prop[shp_order, ]

cat("Shapefile counties matched:", sum(!is.na(shp_order)), "/ 47\n")
if (any(is.na(shp_order))) {
  warning("Unmatched counties: ",
          paste(kenya$COUNTY[is.na(shp_order)], collapse = ", "))
}
```

## Merge Shapefile with Attribute Data

```{r merge-spatial}
# Merge the SpatialPolygonsDataFrame with the reordered analysis data frame.
# The resulting object (kenya.one) carries both geometry and all computed
# SIR/expected-count columns.
kenya.one  <- merge(kenya, data_pop, by.x = "COUNTY", by.y = "County")
data.kenya <- attr(kenya.one, "data")

cat("Merged spatial object has", nrow(data.kenya), "rows\n")
```

---

# Descriptive SIR Maps (Pre-Model)

These choropleth maps display the unadjusted (descriptive) SIR categories
for each year, providing a visual overview of the disease burden distribution
before the Bayesian model is fitted.

## Colour Palette Setup

```{r palette-setup}
# Yellow-Orange-Red palette: low burden (yellow) → high burden (red)
my.palette.yor <- brewer.pal(4, "YlOrRd")

# Suppress axis lines for cleaner map appearance
trellis.par.set(axis.line = list(col = NA))
```

## HIV SIR Maps (2012–2017)

```{r hiv-sir-maps, fig.cap="Figure 1. HIV Standardised Incidence Ratio by county, 2012–2017. Categories: 0.00–0.40 (very low), 0.41–0.70 (low), 0.71–0.99 (moderate), 1.00+ (high burden)."}
# Build one spplot panel per year and arrange in a 2x3 grid
hiv_panels <- lapply(years, function(y) {
  spplot(kenya.one,
         zcol       = paste0("adj_HIV_", y),
         col.regions = my.palette.yor,
         colorkey   = list(space = "right", height = 1.0),
         cuts       = 2,
         main       = as.character(y),
         par.settings = list(fontsize = list(text = 12)))
})

do.call(grid.arrange, c(hiv_panels, list(nrow = 2, ncol = 3)))
```

## TB SIR Maps (2012–2017)

```{r tb-sir-maps, fig.cap="Figure 2. TB Standardised Incidence Ratio by county, 2012–2017."}
tb_panels <- lapply(years, function(y) {
  spplot(kenya.one,
         zcol        = paste0("adj_TB_", y),
         col.regions  = my.palette.yor,
         colorkey    = list(space = "right", height = 1.0),
         cuts        = 2,
         main        = as.character(y),
         par.settings = list(fontsize = list(text = 12)))
})

do.call(grid.arrange, c(tb_panels, list(nrow = 2, ncol = 3)))
```

---

# INLA Model: Bayesian Joint Spatio-Temporal Analysis

## Spatial Adjacency Graph

The spatial neighbourhood structure is derived directly from the county
shapefile using queen-contiguity (shared boundary or vertex).

```{r adjacency-graph}
# Build neighbourhood list from polygon boundaries
adj_graph <- poly2nb(kenya.one)

# Write the INLA-format adjacency file to the working directory
nb2INLA("kenya.graph", adj_graph)
kenya.adj <- file.path(getwd(), "kenya.graph")

cat("Number of counties:", length(adj_graph), "\n")
cat("Average number of neighbours:",
    round(mean(sapply(adj_graph, length)), 2), "\n")
```

## Reshape Data to Long Format

INLA requires the data in **long format**: one row per county × year × disease
combination (47 × 6 × 2 = 564 rows total).

```{r reshape-long}
# ── Step 1: Extract vectors from data_pop by column position ─────────────────
# Column positions correspond to the ordering established when building data_prop.
# HIV observed/expected: columns 3–8 and 17–22 (years 2012–2017)
# TB  observed/expected: columns 10–15 and 24–29
# SIR values:           columns 31–36 (HIV), 38–43 (TB)
observed.HIV  <- as.vector(as.matrix(data_pop[, paste0("obs.HIV", years)]))
observed.TB   <- as.vector(as.matrix(data_pop[, paste0("obs.TB",  years)]))
expected.HIV.v <- as.vector(as.matrix(data_pop[, paste0("exp.HIV", years)]))
expected.TB.v  <- as.vector(as.matrix(data_pop[, paste0("exp.TB",  years)]))
adj.HIV.rate  <- as.vector(as.matrix(data_pop[, paste0("adj.HIV.", years)]))
adj.TB.rate   <- as.vector(as.matrix(data_pop[, paste0("adj.TB.",  years)]))

# ── Step 2: Build the county × year joint data frame ─────────────────────────
n_counties <- nrow(data_pop)   # 47

Year.vec   <- rep(years, each = n_counties)   # 282 elements
County.vec <- as.factor(rep(data_pop[, 1], length(years)))

Joint_data <- data.frame(
  County.ID = as.numeric(County.vec),
  Year.ID   = Year.vec,
  O.HIV     = observed.HIV,
  O.TB      = observed.TB,
  E.HIV     = expected.HIV.v,
  E.TB      = expected.TB.v,
  SIR.HIV   = adj.HIV.rate,
  SIR.TB    = adj.TB.rate
)

# ── Step 3: Stack HIV and TB rows (564-row data frame) ────────────────────────
n.st <- nrow(Joint_data)   # 282 county-year units per disease

data <- data.frame(
  County.ID      = rep(Joint_data$County.ID, 2),
  Year           = rep(Joint_data$Year.ID,   2),
  Year.ID        = rep(Joint_data$Year.ID,   2),
  County.Year.ID = rep(seq_len(n.st),        2)
)

data$Expected <- with(Joint_data, c(E.HIV, E.TB))
data$Observed <- with(Joint_data, c(O.HIV, O.TB))
data$Disease  <- rep(c("HIV", "TB"), each = n.st)

cat("Long-format data: ", nrow(data), "rows x", ncol(data), "columns\n")
```

## Index Construction

INLA identifies random effects through integer indices. Disease-specific effects
use NA for the other disease's rows; shared effects use a single index for both.

```{r build-indices}
# ── Spatial index (1–47, recycled across years) ───────────────────────────────
data$sp.idx <- rep(1:n_counties, length(years) * 2)

# ── Temporal index (1 = 2012, ..., 6 = 2017) ─────────────────────────────────
data$tm.idx <- data$Year - 2011

# ── Disease index (1 = HIV, 2 = TB) ──────────────────────────────────────────
data$dis.idx  <- rep(1:2, each = n.st)
data$intercept <- as.factor(data$dis.idx)   # Disease-specific intercepts

# ── Space-time interaction index ──────────────────────────────────────────────
data$sp.tm.idx <- rep(seq_len(n.st), 2)

# ── Shared (dummy) indices — active for ALL rows ──────────────────────────────
# sp.dum and tm.dum are left NA here; they will be filled via the copy mechanism
data$sp.dum    <- NA                         # Shared spatial effect
data$tm.dum    <- NA                         # Shared temporal effect
data$sp.tm.dum <- data$County.Year.ID        # Shared space-time interaction

# ── Disease-specific spatial indices (NA for the other disease) ───────────────
data$sp.1  <- ifelse(data$dis.idx == 1, data$sp.idx, NA)   # HIV spatial
data$sp.2  <- ifelse(data$dis.idx == 2, data$sp.idx, NA)   # TB  spatial

# Alias: INLA formula references sp.idx1 / sp.idx2
data$sp.idx1 <- data$sp.1
data$sp.idx2 <- data$sp.2

# ── Disease-specific temporal indices ─────────────────────────────────────────
data$tm.1  <- ifelse(data$dis.idx == 1, data$tm.idx, NA)   # HIV temporal
data$tm.2  <- ifelse(data$dis.idx == 2, data$tm.idx, NA)   # TB  temporal

data$tm.idx1 <- data$tm.1
data$tm.idx2 <- data$tm.2

# ── Disease-specific space-time indices ───────────────────────────────────────
data$sp.tm.1  <- ifelse(data$dis.idx == 1, data$sp.tm.idx, NA)
data$sp.tm.2  <- ifelse(data$dis.idx == 2, data$sp.tm.idx, NA)

data$sp.tm.idx1 <- data$sp.tm.1
data$sp.tm.idx2 <- data$sp.tm.2

cat("Index summary (first 6 rows):\n")
print(head(data[, c("County.ID","Year","Disease","dis.idx","sp.idx",
                    "tm.idx","sp.tm.idx")]))
```

## Adjacency Matrices

```{r adjacency-matrices}
# ── Spatial adjacency (47 x 47 binary queen-contiguity matrix) ────────────────
w.sp <- as(nb2mat(adj_graph, style = "B", zero.policy = TRUE), "Matrix")

# ── Temporal adjacency (6 x 6 first-order random-walk structure) ──────────────
# Each year connected only to its immediate predecessor and successor
n_years <- length(years)   # 6
w.tm    <- Diagonal(n_years, x = 0)   # Start with zero diagonal

# Set off-diagonal neighbours
w.tm[1, 2]           <- 1   # 2012 -> 2013
w.tm[n_years, n_years - 1] <- 1   # 2017 -> 2016

for (i in 2:(n_years - 1)) {
  w.tm[i, i - 1] <- 1
  w.tm[i, i + 1] <- 1
}

cat("Spatial adjacency matrix: ", nrow(w.sp), "x", ncol(w.sp), "\n")
cat("Temporal adjacency matrix:", nrow(w.tm), "x", ncol(w.tm), "\n")
```

## Prior Specifications

```{r priors}
# ── Log-Gamma prior on all precision hyperparameters ─────────────────────────
# Shape = 0.5, rate = 0.0005 → weakly informative, allows large variances.
# Applied to: spatial (Besag), temporal (Besag), space-time (IID) effects.
prior.prec <- list(prior   = "loggamma",
                   param   = c(0.5, 0.0005),
                   initial = 0)

# ── Normal prior on copy-effect scaling coefficients (beta) ──────────────────
# Mean 0, precision 1/5.9 ≈ 0.169 (sd ≈ 2.43).
# Range constrained to (0, Inf) to ensure positive association between diseases.
prior.beta.s   <- list(prior = "normal", param = c(0, 1/5.9),
                       fixed = FALSE, initial = 0.01)   # Spatial copy beta
prior.beta.t   <- list(prior = "normal", param = c(0, 1/5.9),
                       fixed = FALSE, initial = 0.01)   # Temporal copy beta
prior.beta.s.t <- list(prior = "normal", param = c(0, 1/5.9),
                       fixed = FALSE, initial = 0.01)   # Space-time copy beta

# Model scaling flag (FALSE = do not scale models to unit variance)
inla.scale <- FALSE
```

## Model Formula

The model is a **shared-component spatio-temporal Poisson model** with:

- Disease-specific spatial effects (Besag intrinsic CAR)
- A shared spatial effect copied to each disease with a scaling beta
- Disease-specific temporal effects (Besag on the first-order RW graph)
- A shared temporal effect with disease-specific copy betas
- Disease-specific space-time IID interactions
- A shared space-time IID interaction with copy betas

```{r model-formula}
formula <- Observed ~ -1 + intercept +

  # ── Disease-specific spatial effects (intrinsic CAR) ─────────────────────
  f(sp.idx1, model = "besag", graph = w.sp,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +
  f(sp.idx2, model = "besag", graph = w.sp,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +

  # ── Shared spatial effect + disease-specific copies ───────────────────────
  f(sp.dum,  model = "besag", graph = w.sp,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +
  f(sp.1, copy = "sp.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s)) +
  f(sp.2, copy = "sp.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s)) +

  # ── Disease-specific temporal effects (first-order RW) ────────────────────
  f(tm.idx1, model = "besag", graph = w.tm,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +
  f(tm.idx2, model = "besag", graph = w.tm,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +

  # ── Shared temporal effect + disease-specific copies ─────────────────────
  f(tm.dum,  model = "besag", graph = w.tm,
    scale.model = inla.scale, hyper = list(prec = prior.prec)) +
  f(tm.1, copy = "tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.t)) +
  f(tm.2, copy = "tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.t)) +

  # ── Disease-specific space-time IID interactions ──────────────────────────
  f(sp.tm.idx1, model = "iid", hyper = list(prec = prior.prec)) +
  f(sp.tm.idx2, model = "iid", hyper = list(prec = prior.prec)) +

  # ── Shared space-time IID interaction + disease-specific copies ───────────
  f(sp.tm.dum,  model = "iid", hyper = list(prec = prior.prec)) +
  f(sp.tm.1, copy = "sp.tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s.t)) +
  f(sp.tm.2, copy = "sp.tm.dum", range = c(0, Inf),
    hyper = list(beta = prior.beta.s.t))
```

## Model Fitting

```{r fit-model, cache=TRUE}
# Fit the Bayesian joint model using INLA.
# cache=TRUE means this chunk is only re-run when its code or dependencies change.
#
# control.compute: request DIC (model comparison), CPO (leave-one-out), and
#                  full config (needed for marginal sampling).
# control.inla:    Laplace strategy with 21 integration points for accuracy.

model.inla <- inla(
  formula,
  family            = "poisson",
  data              = data,
  E                 = Expected,
  control.predictor = list(compute = TRUE),
  verbose           = FALSE,
  control.compute   = list(config = TRUE, dic = TRUE, cpo = TRUE),
  control.inla      = list(strategy = "laplace", npoints = 21)
)

# Save fitted model to disk so it does not need to be re-run each session
save(model.inla, file = "st_model_GAMMA.RData")

# Print the INLA model summary
summary(model.inla)
```

---

# Results: Spatial Effects

## Extract Posterior Means

```{r extract-spatial}
# Exponentiate posterior means to obtain relative risks (RR > 1 = elevated risk)
kenya.one$SHARED_C    <- exp(model.inla$summary.random$sp.dum[,   "mean"])
kenya.one$HIV_C       <- exp(model.inla$summary.random$sp.idx1[,  "mean"])
kenya.one$TB_C        <- exp(model.inla$summary.random$sp.idx2[,  "mean"])

# Combined spatial effect = disease-specific + scaled shared component
kenya.one$Total.HIV_C <- exp(model.inla$summary.random$sp.idx1[, "mean"] +
                              model.inla$summary.random$sp.1[,    "mean"])
kenya.one$Total.TB_C  <- exp(model.inla$summary.random$sp.idx2[, "mean"] +
                              model.inla$summary.random$sp.2[,    "mean"])
```

## Classify Spatial RRs into Categories

```{r classify-spatial}
# Breakpoints for spatial RR maps (note: upper bound is 1.00, not 0.90)
spatial.breaks <- c(0, 0.40, 0.70, 1.00, 6)
spatial.labels <- c("0.00-0.40", "0.41-0.70", "0.71-1.00", ">1.00")

classify_rr <- function(x) {
  cut(x, breaks = spatial.breaks, include.lowest = TRUE, labels = spatial.labels)
}

kenya.one$SHARED   <- classify_rr(kenya.one$SHARED_C)
kenya.one$HIV      <- classify_rr(kenya.one$HIV_C)
kenya.one$TB       <- classify_rr(kenya.one$TB_C)
kenya.one$Total.HIV <- classify_rr(kenya.one$Total.HIV_C)
kenya.one$Total.TB  <- classify_rr(kenya.one$Total.TB_C)
```

## Disease-Specific and Shared Spatial Effects

```{r spatial-effects-map, fig.cap="Figure 3. Posterior mean spatial relative risks: shared component (left), HIV-specific (centre), TB-specific (right)."}
grid.arrange(
  spplot(kenya.one, zcol = "SHARED", col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 2,
         main = "Shared", par.settings = list(fontsize = list(text = 12))),
  spplot(kenya.one, zcol = "HIV",    col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 2,
         main = "HIV",    par.settings = list(fontsize = list(text = 12))),
  spplot(kenya.one, zcol = "TB",     col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 2,
         main = "TB",     par.settings = list(fontsize = list(text = 12))),
  ncol = 3
)
```

## Combined (Total) Spatial Patterns

```{r total-spatial-map, fig.cap="Figure 4. Combined county-level spatial relative risks (disease-specific + shared components): HIV (left) and TB (right)."}
grid.arrange(
  spplot(kenya.one, zcol = "Total.HIV", col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 6,
         main = "HIV: Combined Spatial Pattern",
         par.settings = list(fontsize = list(text = 12))),
  spplot(kenya.one, zcol = "Total.TB",  col.regions = my.palette.yor,
         colorkey = list(space = "right", height = 1.0), cuts = 5,
         main = "TB: Combined Spatial Pattern",
         par.settings = list(fontsize = list(text = 12))),
  ncol = 2
)
```

---

# Results: Temporal Trends

Temporal trends are displayed on the relative risk (RR) scale with 95%
credible intervals (dashed lines).

```{r temporal-trends, fig.cap="Figure 5. Temporal relative risks, 2012–2017. (a) Shared trend; (b) HIV-specific; (c) TB-specific; (d) Combined HIV and TB trends.", fig.height=8}
# Helper: extract and exponentiate temporal posteriors
get_temporal <- function(effect_name) {
  sm <- model.inla$summary.random[[effect_name]]
  list(
    mean  = exp(sm[, "mean"]),
    lower = exp(sm[, "0.025quant"]),
    upper = exp(sm[, "0.975quant"])
  )
}

shared_t  <- get_temporal("tm.dum")
hiv_t     <- get_temporal("tm.idx1")
tb_t      <- get_temporal("tm.idx2")

# Combined trends: disease-specific + shared copy
hiv_copy  <- model.inla$summary.random$tm.1
tb_copy   <- model.inla$summary.random$tm.2

hiv_comb <- list(
  mean  = exp(model.inla$summary.random$tm.idx1[, "mean"]      + hiv_copy[, "mean"]),
  lower = exp(model.inla$summary.random$tm.idx1[, "0.025quant"] + hiv_copy[, "0.025quant"]),
  upper = exp(model.inla$summary.random$tm.idx1[, "0.975quant"] + hiv_copy[, "0.975quant"])
)
tb_comb <- list(
  mean  = exp(model.inla$summary.random$tm.idx2[, "mean"]      + tb_copy[, "mean"]),
  lower = exp(model.inla$summary.random$tm.idx2[, "0.025quant"] + tb_copy[, "0.025quant"]),
  upper = exp(model.inla$summary.random$tm.idx2[, "0.975quant"] + tb_copy[, "0.975quant"])
)

# Helper: draw one temporal trend panel
plot_trend <- function(y_mean, y_lower, y_upper,
                       ylim, main_title, panel_label,
                       line_col = "black") {
  plot(years, y_mean,
       ylim  = ylim,  type  = "l",
       main  = main_title, xlab = panel_label, ylab = "",
       frame = FALSE, col  = line_col, lty = 1, lwd = 2,
       cex.lab = 1.2)
  title(ylab = "Relative Risk", line = 1.9, cex.lab = 1.2)
  lines(years, y_lower, col = line_col, lty = 3, lwd = 0.5)
  lines(years, y_upper, col = line_col, lty = 3, lwd = 0.5)
}

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# (a) Shared temporal trend
plot_trend(shared_t$mean, shared_t$lower, shared_t$upper,
           ylim = c(0.0, 2.0),
           main_title = "Shared Temporal Trend", panel_label = "(a)")

# (b) HIV-specific temporal trend
plot_trend(hiv_t$mean, hiv_t$lower, hiv_t$upper,
           ylim = c(0.8, 1.2),
           main_title = "HIV Temporal Trend", panel_label = "(b)",
           line_col = "red")

# (c) TB-specific temporal trend
plot_trend(tb_t$mean, tb_t$lower, tb_t$upper,
           ylim = c(0.8, 1.2),
           main_title = "TB Temporal Trend", panel_label = "(c)",
           line_col = "blue")

# (d) Combined temporal trends (HIV and TB overlaid)
plot_trend(hiv_comb$mean, hiv_comb$lower, hiv_comb$upper,
           ylim = c(0.5, 1.6),
           main_title = "Combined Temporal Trend", panel_label = "(d)",
           line_col = "red")
lines(years, tb_comb$mean,  col = "blue", lty = 1, lwd = 2)
lines(years, tb_comb$lower, col = "blue", lty = 3, lwd = 0.5)
lines(years, tb_comb$upper, col = "blue", lty = 3, lwd = 0.5)
legend(2012, 1.6, legend = c("HIV", "TB"),
       col = c("red", "blue"), lwd = 2, lty = 1, cex = 0.8)

par(mfrow = c(1, 1))   # Reset layout
```

---

# Results: Space-Time Interaction

The shared space-time interaction captures residual county-year heterogeneity
not explained by the main spatial or temporal effects.

## Extract Posterior Means

```{r space-time-posterior}
# Posterior means of the shared space-time random effect (delta)
# Returns a 47 x 6 matrix (counties x years), filled column-by-column.
delta_s_vec <- model.inla$summary.random$sp.tm.dum[, "mean"]

delta.matrix_s         <- matrix(delta_s_vec, nrow = n_counties, ncol = length(years),
                                 byrow = FALSE)
rownames(delta.matrix_s) <- data.kenya$COUNTY
colnames(delta.matrix_s) <- years

cat("Space-time interaction matrix (first 5 counties):\n")
print(round(delta.matrix_s[1:5, ], 4))
```

## Exceedance Probabilities P(δ > 0 | data)

For each county-year cell, we compute the posterior probability that the
shared space-time interaction is positive (i.e. RR > 1), providing a
formal measure of excess risk.

```{r exceedance-prob}
# Compute P(delta_st > 0 | data) for each county-year combination.
# inla.pmarginal returns P(delta <= 0), so we take 1 - P(delta <= 0).
threshold <- 0   # H0: no excess risk (delta = 0 on log scale)

pp.delta_s <- sapply(
  model.inla$marginals.random[[9]],   # [[9]] = sp.tm.dum marginals
  function(marg) 1 - inla.pmarginal(threshold, marg)
)

# Reshape into county x year matrix and classify into four probability bands
pp.delta.matrix_s <- matrix(pp.delta_s, nrow = n_counties, ncol = length(years),
                             byrow = FALSE)

pp.breaks <- c(0, 0.25, 0.50, 0.75, Inf)
pp.labels <- c("0.00-0.25", "0.26-0.50", "0.51-0.75", "0.76-1.00")

# Build a data frame with one column per year (for merging with spatial object)
pp.delta.factor_s <- data.frame(County = data.kenya$COUNTY)

for (i in seq_along(years)) {
  pp.col <- cut(pp.delta.matrix_s[, i],
                breaks = pp.breaks, include.lowest = TRUE, labels = pp.labels)
  pp.delta.factor_s[[paste0("X2", i + 11)]] <- pp.col   # X212 ... X217
}

cat("Exceedance probability data frame: ",
    nrow(pp.delta.factor_s), "rows,", ncol(pp.delta.factor_s), "columns\n")
```

## Space-Time Interaction Maps

```{r space-time-maps, fig.cap="Figure 6. Posterior probability P(shared space-time interaction > 0 | data) by county, 2012–2017. Values > 0.75 indicate strong evidence of excess county-year risk."}
# Attach exceedance probability columns to the spatial object for plotting
attr(kenya, "data") <- data.frame(data.kenya, pp.delta.factor_s)

my.palette.ygn <- brewer.pal(4, "YlGn")

pp_panels <- lapply(seq_along(years), function(i) {
  zcol <- paste0("X2", i + 11)   # X212, X213, ..., X217
  spplot(obj = kenya, zcol = zcol,
         col.regions = my.palette.ygn, cuts = 2,
         colorkey    = list(space = "right", height = 1.0),
         main        = as.character(years[i]),
         par.settings = list(fontsize = list(text = 12)))
})

do.call(grid.arrange, c(pp_panels, list(nrow = 2, ncol = 3)))
```

---

# Session Information

```{r session-info}
sessionInfo()
```
